---
title: "Time Series Analysis Sheet 5"
date: "9 May 2022"
output:
  pdf_document: default
  html_document: default
---

Tom Kniepkamp; 229134,

Sultan Mahmud Chomon; 230668,

Sk Tanzer Ahmed Siddique; 230764.

<div style="text-align: justify">

## Problem 14 (Linear Filter)

See scan attached.

```{r, include = F}
# Clear Workspace
rm(list = ls())
# Load packages
pacman:: p_load(ZIM, tseries, stats, TSA)
```


## Problem 15 (Combining Linear Filters)

#### Define and test functions: 

- $B^k{y_t}  = y_{t-k},$ for $k \in Z$;

- $V  = 0.2 B^{-1} + 0.6 + 0.2 B$;

- $W  = \Delta_5 = 1- B^5$


```{r}
# define a function for the backshift operator which takes also negative values
B <- function(y, k){
  if (k >= 0){
    z <- rep(NaN, rep = k)
    for (t in (k+1):length(y)){
      z[t] <- y[t - k]
    }
  }
  else {
    z <- c()
    for (t in 1:length(y + k)){
      z <- append(z, y[t - k])
    }
  }
  return(z)
}

# Define V
V <- function(y){
  return(0.2 * B(y, k = -1) + 0.6 + 0.2 * B(y, k = 1))
}

# Define W
W <- function(y){
  return(1- B(y, k = 5))
}
```


### a)

*What could be the use of applying each of the two filters to a time series* ($y_t: t \in \Bbb{Z}$)?

- The first one is a symmetric filter, it can be used to re-scale the time- series. Assuming that the sequence is infinite, no data will be lost in the process (unlike in the empirical example in the code above). 

- The second filter can be used to filter out seasonal differences of a frequency of five. 

### b) 

According to *Proposition 7.3*, we expect the successive application (i.e. convolution) of two linear filters
to be (among other features) *commutative* which means that the order of of application of
successively applied filters does not make a difference to the final result.

In order to determe whether it makes a difference to the end result of applying filter
$V$ or $W$ first, both will be applied to the *airline passengers* sequentially and the results are compared. 

```{r}
AP     <- AirPassengers

# Convolution 1: V * W (Red Line)
conv_1 <- V(AP)
conv_1 <- W(conv_1)

# Convolution 2: W * V (Blue Line)
conv_2 <- W(AP)
conv_2 <- V(conv_2)

# Plot the data 
plot(conv_1, type = "l", col = "red",
     main = "Convolution of Linear Filters",
     xlab = "Time", ylab = "Convulution output") + 
     lines(conv_2, col = "blue", lty = 2)
```

As shown in the graph above, the order of applying the linear filters yields the
same result. Thus it is irrelevant which filter is applied first. these findings 
reinforce the principle of commutativity stated above. 

### c)

Using the results from part b), we define $U = V \circ W = W \circ V$


### d) *Find a linear filter * $\tilde{W}$ *such that* $W = \Delta \circ \tilde{W}$ 

Define $S = \Delta^{-1} = 1 + B + B^2 +\dots$ as the inverse of $\Delta$ such that 
$S \circ \Delta = 1$. Then, we have 

\begin{align}
 W  & = \Delta \circ \tilde{W}\\
 \therefore S \circ W &  = S \circ \Delta \circ \tilde{W}\\
 = S \circ W &  = \tilde{W}\\
\end{align}

## Problem 16

See scan attached. 

## Problem 17 

The monthly time series seems to follow a linear trend (see figure below) and yearly seasonal patterns.
Periodic fluctuations can clearly be seen in the years 1993 - 1997; between 1998 and 2001, these seasonalities cannot be seen as clearly by looking at the graph. 

The figure below shows the dataset with a lnear regression line to visualise the linear trend. 

```{r}
# load data and create TS for monthly data
df <- read.table(file = "Transport.txt")
df <- ts(data = df, start = 1993,  deltat = 1/12, names = "Transport")

# Plot the time series wit a linear OLS fit line to show the linear trend. 
t <- time(df) 
reg_0 <- lm(df ~ t )
plot(df, type = "l", main = "Plot of the Time Series")
points(t, predict.lm(reg_0), type = 'l', col = 'blue', lty = 2)
```

### Eliminate Trend with a High- Pass Filter

From slides chapter 7, slide 4/40: 

Eliminate the trend by successively differencing global polynomial trends. 
As we have established above, the the prices most likely follow a linear trend. 

Apllying theorem 7.1: 

$y_t = m_t + u_t,\, m_t = \beta_0 + \beta_1t^p$, where $p = 1$. 

Thus $\Delta^{p+1}_t m_t = 0 = \Delta_t^2$.

We thus choose a high pass filter of 


$\Delta^r = \Delta^{p+1} = \Delta^2 = (1- B)^2$

The thus detrended time series is shown in the figure below. 

### Eliminate seasonality with a Low- Pass Filter

We have monthly data over several years. There is a price increase every January, 
thus we assume that the prices follow a yearly seasonal pattern. Thus the period for a
season is $l = 12$ months. The low pass filter differences the time series with 
a lag of $l$ and is thus defined as follows: 

$\Delta_l = \Delta_{12}= 1- B_{12}$. 

```{r}
df_diff <- diff(df, differences = 2) # first difference to get rid of trend
df_diff_2 <- diff(df_diff, lag = 12) # get rid of yearly fluctuations
par(mfrow=c(2,2))
plot(df_diff, main = "First- Differenced TS")
spectrum(df_diff)
plot(df_diff_2, main = "De- Seasonalised TS")
spectrum(df_diff_2)
```

### Interpretation of Results

The set of plots above shows the first- differenced plot and the de- seasonalised plot
with their respective periodograms. The first- differenced plot is detrended and clearly
shows seasonality. At the beginning of each year, there is a sudden spike as the transport 
prices are raised in the first month of every new year. The periodogram shows a 
maximum of about 4, due to the irregularities starting at about 1998, however, due
to the nature of the data we still assume yearly seasonality.

The de- seasonlised data is obtained by differencing with a lag of 12, 
accounting for the 12 months. The resulting plot and periodogram show no strong
seasonal pattern although the periodogram peaks at around 3.5.  











