---
title: "Time Series Analysis Sheet 5"
author: "Tom Kniepkamp"
date: "5 May 2022"
output: html_document
---

Tom Kniepkamp; 229134,

Sultan Mahmud Chomon; 230668,

Sk Tanzer Ahmed Siddique; [ADD STUDENT NUMBER].

<div style="text-align: justify">

## Problem 14 (Linear Filter)

### 1. *In what sense is a linear filter linear?*

### 2. 

### 3. What is the advantage of a causal filter over a non- causal filter

- Causal filters have real- time applications while non- causal filters cannot be observed using real data.

- Non- causal filters possibly use data from future time periods which are not observable. 

```{r, include = F}
# Clear Workspace
rm(list = ls())
# Load packages
pacman:: p_load(ZIM, tseries, stats, TSA)
```


## Problem 15 (Combining Linear Filters)

### a)

#### Define and test functions: 

- $B^k{y_t}  = y_{t-k},$ for $k \in Z$;

- $V  = 0.2 B^{-1} + 0.6 + 0.2 B$;

- $W  = \Delta_5 = 1- B^5$


```{r}
# define a function for the backshift operator which takes also negative values
B <- function(y, k){
  if (k >= 0){
    z <- rep(NaN, rep = k)
    for (t in (k+1):length(y)){
      z[t] <- y[t - k]
    }
  }
  else {
    z <- c()
    for (t in 1:length(y + k)){
      z <- append(z, y[t - k])
    }
  }
  return(z)
}

# Define V
V <- function(y){
  return(0.2 * B(y, k = -1) + 0.6 + 0.2 * B(y, k = 1))
}

# Define W
W <- function(y){
  return(1- B(y, k = 5))
}
```


### a)

*What could be the use of applying each of the two filters to a time series* ($y_t: t \in \Bbb{Z}$)?

- The first one is a symmetric filter, it can be used to re-scale the time- series. Assuming that the sequence is infinite, no data will be lost in the process (unlike in the empirical example in the code above). 

- The second filter can be used to filter out seasonal differences of a frequency of five. 

### b) 

According to *Proposition 7.3*, we expect the successive application (i.e. convolution) of two linear filters
to be (among other features) *commutative* which means that the order of of application of
successively applied filters does not make a difference to the final result.

In order to determe whether it makes a difference to the end result of applying filter
$V$ or $W$ first, both will be applied to the *airline passengers* sequentially and the results are compared. 

```{r}
AP     <- AirPassengers

# Convolution 1: V * W (Red Line)
conv_1 <- V(AP)
conv_1 <- W(conv_1)

# Convolution 2: W * V (Blue Line)
conv_2 <- W(AP)
conv_2 <- V(conv_2)

# Plot the data 
plot(conv_1, type = "l", col = "red",
     main = "Convolution of Linear Filters",
     xlab = "Time", ylab = "Convulution output") + 
     lines(conv_2, col = "blue", lty = 2)
```

As shown in the graph above, the order of applying the linear filters yields the
same result. Thus it is irrelevant which filter is applied first. these findings 
reinforce the principle of commutativity stated above. 

### c)

Using the results from part b), we define $U = V \circ W = W \circ V$


### d) *Find a linear filter * $\tilde{W}$ *such that* $W = \Delta \circ \tilde{W}$ 

Define $S = \Delta^{-1} = 1 + B + B^2 +\dots$ as the inverse of $\Delta$ such that 
$S \circ \Delta = 1$. Then, we have 

\begin{align}
 W  & = \Delta \circ \tilde{W}\\
 \therefore S \circ W &  = S \circ \Delta \circ \tilde{W}\\
 = S \circ W &  = \tilde{W}\\
\end{align}

## Problem 17 

The monthly time series seems to follow a linear trend (see figure below) and 
yearly seaonsal patterns. A linear OLS estimate is applied to estimate the linear trend in price increases. 

OLS: 

$y_t = \hat{\beta}_0 + \hat{\beta}_1 t + \varepsilon$

From slide 5/40, we get $p = 1$ and therefore $\Delta^2 m_t = 0$.

```{r}
# load data and create TS for monthly data
df <- read.table(file = "Transport.txt")
df <- ts(data = df, start = 1993,  deltat = 1/12, names = "Transport")

# Plot the time series wit a linear OLS fit line to show the linear trend. 

t <- time(df) 
reg_0 <- lm(df ~ t )
plot(df, type = "l", main = "Plot of the Time Series")
points(t, predict.lm(reg_0), type = 'l', col = 'blue', lty = 2)
```


Create low- pass filter to filter out seasonal effects. We choose a $\lambda = 1/12$. 

Seasonal differencing $\Delta_l := y_t - y_{t-l},\, l = 12$

```{r}
# use for seasonal filter
d <- function(df, l){
  a <- window(x = df, start = time(df)[1]+l/12)
  for (i in 1:length(a)){
    a[i] = a[i] - df[i]
  }
  return(a)
}

df_diff <- diff(df, differences = 3)

plot(df)
plot(df_diff)
```




